---
title: "VES Report"
output: 
    word_document:
         number_sections: true
params:
  current.ves: NA
---


```{r load_lib,echo=FALSE}
# The `params` object is available in the document.
library(rves)
library(knitr)
library(grid)
library(gridExtra)
library(kableExtra)
```

# Basic information about the sounding

The VES  `r params$current.ves$id[[1]]` dataset contains `r length(params$current.ves$ab2)` measurements as show in the following table:
```{r table_data,echo=FALSE}
spacing <- params$current.ves$ab2
apprho <- params$current.ves$appres
ves.df <- data.frame("AB/2" = spacing, "ApparentResitivity" = apprho)
kable(ves.df, digits = 3, align = c('c','c'), caption = "VES DATA") 
#%>%
#  kable_styling(bootstrap_options = "striped", full_width = F)
```

The min and max values of the spacing are `r min(params$current.ves$ab2)` and `r max(params$current.ves$ab2)`  meters respectively, and some of summary statistics of the apparent resistivity are presented in the following table:
```{r appres_summary,echo=FALSE}
summary(params$current.ves$appres)
```

# Interpretation Results

A plot of current VES with the corresponding interpretation is shown in the following figure:

```{r plotves,echo=FALSE}
plot(params$current.ves, type = "ves")
``` 


```{r def_values,echo=FALSE}
rho.true <- params$current.ves$rhopar
thickness <- params$current.ves$thickpar
interpretation.ves <- cbind(rho.true, thickness)
interpretation.ves.df <- as.data.frame(interpretation.ves)
names(interpretation.ves.df) <- c("TrueResistivity", "Thickness")
nlayers <- length(rho.true)
layers.id <- vector("character", length = nlayers)
for(i in 1:nlayers){
  layers.id[i] <- paste0("Layer", as.character(i))
}
row.names(interpretation.ves.df) <- layers.id
```

The interpreted model is composed of `r nlayers` layers with the following properties:

```{r table,echo=FALSE}
kable(interpretation.ves.df, digits = 3, align = c('r','r'), 
      caption = "ESTIMATED MODEL") #%>%
#  kable_styling(bootstrap_options = "striped", full_width = F)
```

# Model Diagnostic

The quality of the estimated model can be checked using the following figure:

```{r plot_diagnostic,echo=FALSE,message=FALSE,warnings=FALSE}
p1 <- plot(params$current.ves, type = "diagnostic")
grid.draw(p1)
```
