---
title: "VES Smoothing using the rves package"
author: "Oscar Garcia-Cabrejo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rves)
library(ggplot2)
```

# Introduction

Importance of VES for subsurface investigations. Specially for groundwater.

VES measurements of apparent resistivity and node spacing. Values of apparent resistivity sometimes include noise which is an obstacle for a proper parameter estimation.

Noise in VES can be reduced using smoothing techniques. Definition of smoothing. Techniques for smoothing.


# Basic Concepts

## Smoothing via splines

Let $\{x_{i},\rho_{\text{app},i}:i=1,\ldots.n\}$ be a set of measurements of electrode spacing $x$ and apparent resistivity $\rho_{\text{app}}$ in what is called a VES (Vertical Electrical Sounding). It is assumed that the measurements of apparent resistivity are described by a model $\rho_{\text{app},i}=\hat{\rho_{\text{app}},i} + \epsilon_{i}$ where $\epsilon_{i}$ are a set of independent, zero mean with constant variance random variables, and $\hat{\rho}_{\text{app},i}$ is a cubic smoothing spline that approximates the measurements  $\rho_{\text{app},i}$. The function $\hat{\rho}_{\text{app},i}$ can be estimated as the minimum of:
$$
\sum\limits_{i=1}^{n}[\rho_{\text{app},i}-\hat{\rho}_{\text{app},i}]^{2} + \lambda \int \hat{\rho}_{\text{app}}^{''}(x)^{2}dx
$$
where $\lambda$ is the smoothing parameter, and the function $\hat{\rho}_{\text{app},i}$ is assumed as continuous with at least second order derivative. The previous equation aims at defining a function that approximates the measured values (first term) while minimizing the roughness of the estimated function (second term). This tradeoff is controlled by the $\lambda$ parameter that is called the smoothing parameter. If $\lambda \ge 0$  


## Smoothing via Kernel Regression

## Smoothing via Wavelets


# Example

## VES 1

The first step in the interpretation of a VES is to load the __rves__ package: 
```{r load_library,echo=TRUE}
library(rves)
```
For this example, the VES ves\_data1 included in the package is used to test the different functions included in __rves__:
```{r load_data,echo=TRUE}
data("ves_data1")
```
The electrode spacing and the apparent resistivity values are defined:
```{r define_ves1,echo=TRUE}
ab2 <- ves_data1$ab2
apprho <- ves_data1$apprho
```
and these are used to define the VES object inside R:
```{r define_ves2,echo=TRUE}
sev1a <- ves(id= "Sounding 1", ab2 = ab2, apprho = apprho)
```
The plot of the VES used as example can be seen in the following plot:
```{r plot_sev1a, fig.height=4,fig.width=6,fig.align='center'}
p1 <- plot(sev1a)
```


### Smoothing via splines

```{r smoothspline,echo=TRUE}
res.ss <- smoothing_ves(sev1a, method = "smooth.spline")
```

```{r smoothspline_organize,echo=TRUE}
ab2.def <- rep(res.ss$ab2, 2)
apprho.def <- vector('numeric', length = 36)
apprho.def[1:18] <- res.ss$apprho
print(length(sev1a$appres))
apprho.def[19:36] <- sev1a$appres
meas.type <- vector('character', length = 36)
meas.type[1:18] <- rep("smooth.spline", 18)
meas.type[19:36] <- rep("raw", 18)
res.ss.df <- data.frame(ab2 = ab2.def, apprho = apprho.def, 
                        type = as.factor(meas.type))
```

```{r plot_smoothspline,fig.width=6,fig.height=4,fig.align='center'}
p.ss <- ggplot() + geom_point(aes(x = ab2, y = apprho, color = type), 
                              data = res.ss.df) +
  scale_x_log10() + 
  scale_y_log10() + 
  ggtitle('Comparison') + 
  theme_bw()
print(p.ss)  
```

```{r smoothspline_vesdef,echo=TRUE}
sev1ss <- ves(id = "Sounding 1-Smoothing Spline", ab2 = res.ss$ab2, 
              apprho = res.ss$apprho)
```

```{r initial_sol, echo= TRUE}
rho <- c(40,70,30, 20)
thick <- c(2,10,50,500)
par0 <- c(rho, thick)
```

```{r estimate_smoothspline, echo=TRUE}
res.ss.nls <- calibrate_nls(sev1ss, par0, iterations = 10, ireport = 5)
```

```{r assign_results_ss_nls,echo=TRUE}
sev1ss$rhopar <- res.ss.nls$rho
sev1ss$thickpar <- res.ss.nls$thickness
sev1ss$interpreted <- TRUE
```

```{r plot_sevss,fig.width=6,fig.height=4,fig.align='center'}
p3 <- plot(sev1ss)
print(p3)
```

### Kernel Regression

```{r kernelregression,echo=TRUE}
res.kr <- smoothing_ves(sev1a, method = "kernel.regression", bw = 0.5)
```

```{r kernelregression_organize,echo=TRUE}
ab2.def <- rep(res.kr$ab2, 2)
apprho.def <- vector('numeric', length = 36)
apprho.def[1:18] <- res.kr$apprho
apprho.def[19:36] <- sev1a$appres
meas.type <- vector('character', length = 36)
meas.type[1:18] <- rep("kernel.regression", 18)
meas.type[19:36] <- rep("raw", 18)
res.kr.df <- data.frame(ab2 = ab2.def, apprho = apprho.def, 
                        type = as.factor(meas.type))
```

```{r plot_kernelregression,fig.width=6,fig.height=4,fig.align='center'}
p.kr <- ggplot() + geom_point(aes(x = ab2, y = apprho, color = type), 
                              data = res.kr.df) +
  scale_x_log10() + 
  scale_y_log10() + 
  ggtitle('Comparison') +
  theme_bw()
print(p.kr)  
```

```{r kr_vesdef,echo=TRUE}
sev1kr <- ves(id = "Sounding 1-Kernel Regression", ab2 = res.kr$ab2, 
              apprho = res.kr$apprho)
```

```{r estimate_kr, echo=TRUE}
res.kr.nls <- calibrate_nls(sev1kr, par0, iterations = 10, ireport = 5)
```

```{r assign_results_kr_nls,echo=TRUE}
sev1kr$rhopar <- res.kr.nls$rho
sev1kr$thickpar <- res.kr.nls$thickness
sev1kr$interpreted <- TRUE
```

```{r plot_sevkr,fig.width=6,fig.height=4,fig.align='center'}
p4 <- plot(sev1kr)
print(p4)
```

### Wavelet Smoothing


```{r wave,echo=TRUE}
#library(wavethresh)
#print(names(sev1a))
#ab2.out <- pumpingtest::logseq(min(log10(sev1a$ab2)), max(log10(sev1a$ab2)), 32)
#apprho.approx <- approx(log10(sev1a$ab2),log10(sev1a$appres),
#                        xout = log10(ab2.out), method = 'linear')
#waveletwmap <- wd(apprho.approx$y, family="DaubLeAsymm", filter.number=10)
#softthreshwmap <- threshold(waveletwmap, type="soft", policy="universal")
#hardthreshwmap <- threshold(waveletwmap, type="hard", policy="universal")
#s.soft <- wr(softthreshwmap)
#s.hard <- wr(hardthreshwmap)
```

```{r wave1,fig.height=4,fig.width=6,fig.align='center'}
#plot(sev1a$ab2, sev1a$appres, type = "p", log = 'xy', main = 'Wavelet Smoothing', 
#     pch = 16)
#points(ab2.out, 10^s.soft, col = 'red')
#points(ab2.out, 10^s.hard, col = 'blue')
```

```{r wave2,echo=TRUE}
#soft <- pracma::interp1(ab2.out, s.soft, xi = c(3.1, sev1a$ab2[2:17], 299.9))
#hard <- pracma::interp1(ab2.out, s.hard, xi = c(3.1, sev1a$ab2[2:17], 299.9))
#ws.df <- data.frame(ab2 = sev1a$ab2, raw = sev1a$appres, soft = 10^soft, 
#                    hard = 10^hard)
```

```{r wave3,fig.height=4,fig.width=6,fig.align='center'}
#p1 <- ggplot() + geom_point(aes(x = ab2, y = raw), data = ws.df) + 
#  geom_point(aes(x = ab2, y = soft), data = ws.df, col = "blue") +
#  geom_point(aes(x = ab2, y = hard), data = ws.df, col = "red") +
#  theme_bw() + 
#  ggtitle("Wavelet Smoothing") +
#  scale_x_log10() + 
#  scale_y_log10() +
#  ylab('Apparent Resistivity') +
#  xlab("AB2(m)")
#print(p1)
```
