---
title: "VES Smoothing using the rves package"
author: "Oscar Garcia-Cabrejo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rves)
library(ggplot2)
```

# Introduction

Vertical Electrical Sounding (VES) is one of the oldest resistivity methods and its the main geophysical method used in groundwater exploration  due to its cost and investigation depth. In a VES, measurements of apparent resistivity are taked for different values of node spacing. As in any measurement of a physical property, the values of apparent resistivity sometimes include some degree of noise which is an obstacle for a proper parameter estimation (thicknesses and real resistivities). Given the ubiquity of noise in measurements of the apparent resistivities, there are no methodologies to remove the noise in VES. 

Noise in VES can be reduced using smoothing techniques. Definition of smoothing. Techniques for smoothing.


# Basic Concepts

## Smoothing via splines

Let $\{x_{i},\rho_{\text{app},i}:i=1,\ldots.n\}$ be a set of measurements of electrode spacing $x$ and apparent resistivity $\rho_{\text{app}}$ in what is called a VES (Vertical Electrical Sounding). It is assumed that the measurements of apparent resistivity are described by a model $\rho_{\text{app},i}=\hat{\rho}_{\text{app},i} + \epsilon_{i}$ where $\epsilon_{i}$ are a set of independent random variables with  zero mean and constant variance, and $\hat{\rho}_{\text{app},i}$ is a cubic smoothing spline that approximates the measurements  $\rho_{\text{app},i}$. This cubic spline is a third order polynomial defined as:
$$
\hat{\rho}_{\text{app},i}(x) = a_{i}(x-x_{i})^{3}+b_{i}(x-x_{i})^{2} + c_{i}(x-x_{i}) + d_{i}
$$
where $a_{i},b_{i},c_{i}$ and $d_{i}$ are fixed coefficients that can be estimated from the measured data. The coefficients of the function $\hat{\rho}_{\text{app},i}$ can be estimated as the minimum of:
$$
\sum\limits_{i=1}^{n}[\rho_{\text{app},i}-\hat{\rho}_{\text{app},i}]^{2} + \lambda \int \hat{\rho}_{\text{app}}^{''}(x)^{2}dx
$$
where $\lambda$ is the smoothing parameter, and the function $\hat{\rho}_{\text{app},i}$ is assumed as continuous with at least second order derivative. The previous equation aims at defining a function that approximates the measured values (first term) while minimizing the roughness of the estimated function (second term). This tradeoff is controlled by the $\lambda$ parameter that is called the smoothing parameter. If $\lambda \rightarrow 0$ then there is no smoothing and the spline interpolation is obtained. If $\lambda \rightarrow \infty$ then all the weight is given to the roughness penalty and a least-squares estimate is obtained.    


## Smoothing via Kernel Regression

Another approach to smooth the VES data is to use local polynomial regressions. In this approach, the function that approximates the apparent resistivity measurements is a polynomial of order $p$ defined as:
$$
\hat{\rho}_{\text{app},i}(x) = a_{0} + a_{1}(x-x_{i})+\frac{a_{2}}{2!}(x-x_{i})^{2} + \ldots + \frac{a_{p}}{p!}(x-x_{i})^{p}
$$
where the $(a_{i},i=0,\ldots,p)$ are fixed coefficients that are estimated by minimizing:
$$
\sum_{i=1}^{n}w_{i}(x)[\rho_{\text{app}}(x_{i})-\hat{\rho}_{\text{app},i}(x)]^{2}
$$

The coefficientes $a_{i}$ can be  estimated using weighted linear least-squares:
$$
\boldsymbol{a}=(\boldsymbol{X}_{X}^{T} \boldsymbol{W}_{X} \boldsymbol{X}_{X})^{-1} \boldsymbol{X}_{X}^{T}\boldsymbol{W}_{X}\boldsymbol{Y}
$$
where the matrix $\boldsymbol{X}_{X}$ is defined as:
$$
\boldsymbol{X}_{X}=\left[
\begin{array}{cccc}
1& (x-x_{1})& \cdots& \frac{(x-x_{1})^{p}}{p!}\\
1& (x-x_{2})& \cdots& \frac{(x-x_{2})^{p}}{p!}\\
\vdots& \vdots& \cdots& \vdots\\
1& (x-x_{n})& \cdots& \frac{(x-x_{n})^{p}}{p!}\\
\end{array}
\right]
$$
and the weight matrix is given by:
$$
\boldsymbol{W}_{X}=\left[
\begin{array}{cccc}
w(x_{1})& 0& \cdots & 0 \\
0& w(x_{2})& \cdots & 0 \\
\vdots& \vdots& \vdots& \vdots \\
0& 0& \cdots & w(x_{n})
\end{array}
\right]
$$


## Smoothing via Wavelet Shrinkage

Let $\{x_{i},\rho_{\text{app},i}:i=1,\ldots.n\}$ be a set of measurements of a VES ( electrode spacing $x$ and apparent resistivity $\rho_{\text{app}}$). Again it is assumed that the measurements of the apparent resistivity $\rho_{\text{app},i}^{m}$ are given by:
$$
\rho_{\text{app},i}^{m}(x) = \hat{\rho}_{\text{app},i}(x) + \epsilon(x)  
$$
where $\hat{\rho}_{\text{app},i}(x)$ are the true values of the apparent resistivity and $\epsilon(x)$ is a noise component that needs to be filtered. Let $W[\cdot]$ and $W^{-1}[\cdot]$ be the forward and inverse wavelet tranform operators, and $D[\cdot,\lambda]$ the denoising operator defined in terms of the $\lambda$ parameter. The smoothing procedure using wavelet shrinkage has three steps:

1. Calculate the wavelet transform of the measured apparent resistivity curve:
$$
Y = W[\rho_{\text{app},i}^{m}(x)]
$$
2. Apply the denoising operator to the wavelet transform of the measured resistivity:
$$
Z=D[Y,\lambda]
$$
3. Calculate the inverse wavelet transform of the denoised wavelet transform:
$$
\hat{\rho}_{\text{app},i}(x)= W^{-1}[Z]
$$
As seen in the previous equations, the key of this smoothing technique is the definition of the $\lambda$ parameter called _shinkring parameter_. As its name suggets, the role of $\lambda$ is to shrink the magnitude of the wavelet transform of a signal $U$ using the following rule:  


$$
D[U,\lambda]=\text{sgn}{(U)} \max{(0,|U|-\lambda)}
$$

The action of this operator is to nullify the values of $U$ for which $|U| < \lambda$, and reduce the values of $U$ to $|U|-\lambda$ in the case that $|U| > \lambda$. The small values of $U$ are associated with the high frequency variations in the signals, in other words, this variations are called noise. 

# Example

## VES 1

The first step in the interpretation of a VES is to load the __rves__ package: 
```{r load_library,echo=TRUE}
library(rves)
```
For this example, the VES ves\_data1 included in the package is used to test the different functions included in __rves__:
```{r load_data,echo=TRUE}
data("ves_data1")
```
The electrode spacing and the apparent resistivity values are defined:
```{r define_ves1,echo=TRUE}
ab2 <- ves_data1$ab2
apprho <- ves_data1$apprho
```
and these are used to define the VES object inside R:
```{r define_ves2,echo=TRUE}
sev1a <- ves(id= "VES1", ab2 = ab2, apprho = apprho)
```
The plot of the VES used as example can be seen in the following plot:
```{r plot_sev1a, fig.height=4,fig.width=6,fig.align='center'}
p1 <- plot(sev1a, type = "ves")
```


### Smoothing via splines
The smoothing of the VES1 using smoothing splines is achieved using the _smoothing\_ves_ function:
```{r smoothspline,echo=TRUE}
res.ss <- smoothing_ves(sev1a, method = "smooth.spline")
```
Now the results of the smoothing procedure are extracted and assigned to a data.frame:
```{r smoothspline_organize,echo=TRUE}
ab2.def <- rep(res.ss$ab2, 2)
apprho.def <- vector('numeric', length = 36)
apprho.def[1:18] <- res.ss$apprho
print(length(sev1a$appres))
apprho.def[19:36] <- sev1a$appres
meas.type <- vector('character', length = 36)
meas.type[1:18] <- rep("smooth.spline", 18)
meas.type[19:36] <- rep("raw", 18)
res.ss.df <- data.frame(ab2 = ab2.def, apprho = apprho.def, 
                        type = as.factor(meas.type))
```
A comparison of the measured and smoothed apparent resistivity values are seed in the following figure, where it is evident that there are small differences between the resistivity values. The smoothing procedure using splines corrects the values at intermediate and large spacings. 
```{r plot_smoothspline,fig.width=6,fig.height=4,fig.align='center'}
p.ss <- ggplot() + geom_point(aes(x = ab2, y = apprho, color = type), 
                              data = res.ss.df) +
  scale_x_log10() + 
  scale_y_log10() + 
  ggtitle('Comparison') + 
  theme_bw()
print(p.ss)  
```
Now, the smoothed resistivity values are used to define a new ves object:

```{r smoothspline_vesdef,echo=TRUE}
sev1ss <- ves(id = "Sounding 1-Smoothing Spline", ab2 = res.ss$ab2, 
              apprho = res.ss$apprho)
```
The initial solution for the calibration procedure are defined as follows:
```{r initial_sol, echo= TRUE}
rho <- c(40,70,30, 20)
thick <- c(2,10,50,500)
par0 <- c(rho, thick)
```
and the true resistivities and layer thicknesses are estimated using the conventional nonlinear least-squares approach:
```{r estimate_smoothspline, echo=TRUE}
res.ss.nls <- calibrate_nls(sev1ss, par0, iterations = 10, ireport = 5)
```
The results of the calibration are assigned to the ves object
```{r assign_results_ss_nls,echo=TRUE}
sev1ss$rhopar <- res.ss.nls$rho
sev1ss$thickpar <- res.ss.nls$thickness
sev1ss$interpreted <- TRUE
```
and the results are plotted.
```{r plot_sevss,fig.width=6,fig.height=4,fig.align='center'}
p3 <- plot(sev1ss, type = "ves")
print(p3)
```
The smoothing using splines causes a reduction in the calibration error which now is equal to $0.81$ in contrast to $1.35$ obtained whithout smoothing. 

### Kernel Regression

```{r kernelregression,echo=TRUE}
res.kr <- smoothing_ves(sev1a, method = "kernel.regression", bw = 0.5)
```

```{r kernelregression_organize,echo=TRUE}
ab2.def <- rep(res.kr$ab2, 2)
apprho.def <- vector('numeric', length = 36)
apprho.def[1:18] <- res.kr$apprho
apprho.def[19:36] <- sev1a$appres
meas.type <- vector('character', length = 36)
meas.type[1:18] <- rep("kernel.regression", 18)
meas.type[19:36] <- rep("raw", 18)
res.kr.df <- data.frame(ab2 = ab2.def, apprho = apprho.def, 
                        type = as.factor(meas.type))
```

```{r plot_kernelregression,fig.width=7,fig.height=4,fig.align='center'}
p.kr <- ggplot() + geom_point(aes(x = ab2, y = apprho, color = type), 
                              data = res.kr.df) +
  scale_x_log10() + 
  scale_y_log10() + 
  ggtitle('Comparison') +
  theme_bw()
print(p.kr)  
```

```{r kr_vesdef,echo=TRUE}
sev1kr <- ves(id = "Sounding 1-Kernel Regression", ab2 = res.kr$ab2, 
              apprho = res.kr$apprho)
```

```{r estimate_kr, echo=TRUE}
res.kr.nls <- calibrate_nls(sev1kr, par0, iterations = 10, ireport = 5)
```

```{r assign_results_kr_nls,echo=TRUE}
sev1kr$rhopar <- res.kr.nls$rho
sev1kr$thickpar <- res.kr.nls$thickness
sev1kr$interpreted <- TRUE
```

```{r plot_sevkr,fig.width=6,fig.height=4,fig.align='center'}
p4 <- plot(sev1kr, type = "ves")
print(p4)
```

### Wavelet Smoothing


```{r wave,echo=TRUE}
res.wv <- smoothing_ves(sev1a, method = "wavelet")
```

```{r wave1,echo=TRUE}
ab2.def <- rep(res.wv$ab2, 2)
apprho.def <- vector('numeric', length = 36)
apprho.def[1:18] <- res.wv$apprho
apprho.def[19:36] <- sev1a$appres
meas.type <- vector('character', length = 36)
meas.type[1:18] <- rep("wavelet.thresholding", 18)
meas.type[19:36] <- rep("raw", 18)
res.wv.df <- data.frame(ab2 = ab2.def, apprho = apprho.def, 
                        type = as.factor(meas.type))
```


```{r wave2,fig.height=4,fig.width=6,fig.align='center'}
p.wv <- ggplot() + geom_point(aes(x = ab2, y = apprho, color = type), 
                              data = res.wv.df) +
  scale_x_log10() + 
  scale_y_log10() + 
  ggtitle('Comparison') +
  theme_bw()
print(p.wv)  
```

```{r wave3,echo=TRUE}
sev1wv <- ves(id = "Sounding 1-Wavelet Thresholding", ab2 = res.wv$ab2, 
              apprho = res.wv$apprho)
```

```{r wave4,echo=TRUE}
res.wv.nls <- calibrate_nls(sev1wv, par0, iterations = 10, ireport = 5)
```

```{r wave5,echo=TRUE}
sev1wv$rhopar <- res.wv.nls$rho
sev1wv$thickpar <- res.wv.nls$thickness
sev1wv$interpreted <- TRUE
```



```{r wave6,fig.height=4,fig.width=6,fig.align='center'}
p5 <- plot(sev1wv, type = "ves")
print(p5)
```
